---
title: "Milestone Report"
output: html_document
author: M.Welt
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(stringr) ## read lines, count words
```
```{r Cache_functions, echo=FALSE}
# Function to load saved R objects from a .RData file if the file exists
loadObjects <- function(fileName) {
  # Append ".RData" extension to the filename
  filePath <- file.path("cache", paste0(fileName, ".RData"))
  
  # Check if the file exists before attempting to load
  if (file.exists(filePath)) {
    # Load objects into the global environment
    load(filePath, envir = .GlobalEnv)
  }
}
save_objects <- function(objects, file_name) {
  if (!dir.exists("cache")) {
    dir.create("cache", recursive = TRUE)
  } 
  filePath <- file.path("cache", paste0(file_name, ".RData"))
  message(filePath)

  save(list = objects, file = filePath)
}
# Function to format numbers with thousand separators
format_number <- function(x) {
  if (is.numeric(x)) {
    format(x, big.mark = ",", scientific = FALSE)
  } else {
    x
  }
}

show_table <- function(report) {
  kable(report, align = rep("r", ncol(report))) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  column_spec(1:ncol(report), border_right = "1px solid lightgray")
}
```

## A basic report of summary statistics about the data sets
https://www.coursera.org/learn/data-science-project/supplement/BePVz/task-2-exploratory-data-analysis


```{r basic_report}
dir <- c("Coursera-SwiftKey/final/en_US")
filelist <- list.files(dir)
filelistFullPath <- file.path(dir, filelist)
CACHE_FILE <- "basic_report"
loadObjects(CACHE_FILE)
if (!exists(CACHE_FILE) || is.null(basic_report)) 
{
  basic_report <- lapply(filelistFullPath, function(x) round(file.info(x)[1]/1024/1024))
  basic_report <- as.data.frame(basic_report)
  # Initialize lists for line and word counts
  line_counts <- c()
  word_counts <- c()
  
  # Read first 100 lines and count lines & words in a single pass
  for (file in filelistFullPath) {
    lines <- readLines(file, warn = FALSE)  
    line_counts <- c(line_counts, length(lines))  # Count lines
    word_counts <- c(word_counts, sum(str_count(lines, "\\S+")))  # Count words
  }  

  names(basic_report) <- gsub("\\.txt$", "", filelist)
  basic_report <- rbind(basic_report, line_counts, word_counts)
  basic_report <- as.data.frame(lapply(basic_report, format_number))
  rownames(basic_report) <- c("size(MB)", "Lines", "Words")
  save_objects(c("basic_report"), CACHE_FILE)
}



```
```{r show_basic_report, echo=FALSE}
show_table(basic_report)
```


### Word frequencies

**Some words are more frequent than others - what are the distributions of word frequencies?** 
Read a sample of 10,000 lines from all files
```{r read sample text}
set.seed(12345)
sample_size <- 10000
# Define function to read a random sample of lines
read_random_sample <- function(file_path, sample_size = 10000) {
  # Count total lines in the file
  total_lines <- basic_report["Lines", gsub("\\.txt$", "", basename(file_path))] # Get total lines from `basic_report`
  
  # Convert from character to numeric (since `basic_report` stores formatted numbers)
  total_lines <- as.numeric(gsub(",", "", total_lines))
  
  # Select random line numbers
  sampled_lines <- sort(sample(1:total_lines, min(sample_size, total_lines)))
  
  # Read only the sampled lines
  text_data <- readLines(file_path, warn = FALSE)[sampled_lines]
  
  return(text_data)
}

# Read random lines from all files
text_data <- unlist(lapply(filelistFullPath, read_random_sample))


```
Analyze word frequencies

```{r Analyze word frequencies, message=FALSE, warning=FALSE}
library(tm) #Text Mining Package
CACHE_FILE <- "corpus"
loadObjects(CACHE_FILE)
if (!exists(CACHE_FILE) || is.null(corpus)) 
{
  # Create a Corpus
  corpus <- Corpus(VectorSource(text_data))
  
  # Clean Text Data
  corpus <- tm_map(corpus, content_transformer(tolower)) # Convert to lowercase
  corpus <- tm_map(corpus, removePunctuation)            # Remove punctuation
  corpus <- tm_map(corpus, removeNumbers)               # Remove numbers
  corpus <- tm_map(corpus, removeWords, stopwords("en")) # Remove stopwords
  corpus <- tm_map(corpus, stripWhitespace)             # Remove extra spaces
  
  # Create Term-Document Matrix
  tdm <- TermDocumentMatrix(corpus)
  word_freq <- sort(rowSums(as.matrix(tdm)), decreasing = TRUE)
  word_freq_df <- data.frame(word = names(word_freq), freq = word_freq)
  
  save_objects(c("corpus","tdm","word_freq","word_freq_df"), CACHE_FILE)
}

```
Word Cloud of Most Frequent Words
```{r Word Cloud, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(wordcloud)

wordcloud(words = word_freq_df$word, freq = word_freq_df$freq, min.freq = 10,
          max.words = 100, colors = brewer.pal(8, "Dark2"), random.order = FALSE)


```

Bar Plot of Top 20 Words

```{r Bar Plot of Top 20 Words, echo=FALSE}
# --- PLOT : Bar Plot of Top 20 Words ---
top_n <- 20
ggplot(head(word_freq_df, top_n), aes(x = reorder(word, -freq), y = freq)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(x = "Words", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




**What are the frequencies of 2-grams and 3-grams in the dataset?**
```{r 2-grams and 3-grams, warning=FALSE}
# Convert Corpus to Plain Text
CACHE_FILE <- "ngrams"
loadObjects(CACHE_FILE)
if (!exists("bigram_df") || is.null(bigram_df)) 
{
  text_clean <- sapply(corpus, as.character)
  text_clean <- paste(text_clean, collapse = " ")
  
  nGram <- function (wordsLength) {
    tokenizer <- function(x) {
      unlist(lapply(ngrams(words(x), wordsLength), paste, collapse = " "))
    }
    # Create Term-Document Matrices
    tdm_ngram <- TermDocumentMatrix(VCorpus(VectorSource(text_clean)), control = list(tokenize = tokenizer))
    # Convert Matrices to Data Frames
    ngram_freq <- sort(rowSums(as.matrix(tdm_ngram)), decreasing = TRUE)
    ngram_df <- data.frame(ngram = names(ngram_freq), freq = ngram_freq)
  }
  bigram_df <<- nGram(2)
  trigram_df <<- nGram(3)
  save_objects(c("bigram_df","trigram_df"), CACHE_FILE)
}
drawBars <- function(ds,title) {
  ggplot(head(ds, 20), aes(x = reorder(ngram, -freq), y = freq)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  theme_minimal() +
  labs(title = title, x = "", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
}

drawBars(bigram_df,"Top 20 Most Frequent Bigrams")
drawBars(trigram_df,"Top 20 Most Frequent Trigrams")

drawCloud <- function(ds,title) {
  wordcloud(words = ds$ngram, freq = ds$freq, min.freq = 10,
            max.words = 100, colors = brewer.pal(8, "Dark2"), random.order = FALSE)
}

drawCloud(bigram_df)
drawCloud(trigram_df)
```
```{r Gram charts, echo=FALSE}
# --- PLOT 1: Top 20 Bigrams ---
# ggplot(head(bigram_df, 20), aes(x = reorder(ngram, -freq), y = freq)) +
#   geom_bar(stat = "identity", fill = "steelblue") +
#   theme_minimal() +
#   labs(title = "Top 20 Most Frequent Bigrams", x = "Bigrams", y = "Frequency") +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1))
# 
# # --- PLOT 2: Top 20 Trigrams ---
# ggplot(head(trigram_df, 20), aes(x = reorder(ngram, -freq), y = freq)) +
#   geom_bar(stat = "identity", fill = "steelblue") +
#   theme_minimal() +
#   labs(title = "Top 20 Most Frequent Trigrams", x = "Trigrams", y = "Frequency") +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1))
# 
# # # --- PLOT 3: Word Cloud for Bigrams ---
# wordcloud(words = bigram_df$ngram, freq = bigram_df$freq, min.freq = 10,
#           max.words = 100, colors = brewer.pal(8, "Dark2"), random.order = FALSE)
# # 
# # # --- PLOT 4: Word Cloud for Trigrams ---
# wordcloud(words = trigram_df$ngram, freq = trigram_df$freq, min.freq = 10,
#           max.words = 100, colors = brewer.pal(8, "Set3"), random.order = FALSE)
```



How many unique words do you need in a frequency sorted dictionary to cover 50% of all word instances in the language? 90%? 

How do you evaluate how many of the words come from foreign languages? 

Can you think of a way to increase the coverage -- identifying words that may not be in the corpora or using a smaller number of words in the dictionary to cover the same number of phrases?



## Task 3 - Modeling

https://www.coursera.org/learn/data-science-project/supplement/2IiM9/task-3-modeling


The goal here is to build your first simple model for the relationship between words. This is the first step in building a predictive text mining application. You will explore simple models and discover more complicated modeling techniques.

Tasks to accomplish

Build basic n-gram model - using the exploratory analysis you performed, build a basic 
n-gram model
 for predicting the next word based on the previous 1, 2, or 3 words.

Build a model to handle unseen n-grams - in some cases people will want to type a combination of words that does not appear in the corpora. Build a model to handle cases where a particular n-gram isn't observed.

Questions to consider

How can you efficiently store an n-gram model (think Markov Chains)?

How can you use the knowledge about word frequencies to make your model smaller and more efficient?

How many parameters do you need (i.e. how big is n in your n-gram model)?

Can you think of simple ways to "smooth" the probabilities (think about giving all n-grams a non-zero probability even if they aren't observed in the data) ?

How do you evaluate whether your model is any good?

How can you use 
backoff models
 to estimate the probability of unobserved n-grams?

Hints, tips, and tricks

As you develop your prediction model, two key aspects that you will have to keep in mind are the size and runtime of the algorithm. These are defined as:

Size: the amount of memory (physical RAM) required to run the model in R

Runtime: The amount of time the algorithm takes to make a prediction given the acceptable input

Your goal for this prediction model is to minimize both the size and runtime of the model in order to provide a reasonable experience to the user.

Keep in mind that currently available predictive text models can run on mobile phones, which typically have limited memory and processing power compared to desktop computers. Therefore, you should consider very carefully (1) how much memory is being used by the objects in your workspace; and (2) how much time it is taking to run your model. Ultimately, your model will need to run in a Shiny app that runs on the 
shinyapps.io
 server.

Tips, tricks, and hints

Here are a few tools that may be of use to you as you work on their algorithm:

object.size(): this function reports the number of bytes that an R object occupies in memory

Rprof(): this function runs the profiler in R that can be used to determine where bottlenecks in your function may exist. The profr package (available on CRAN) provides some additional tools for visualizing and summarizing profiling data.

gc(): this function runs the garbage collector to retrieve unused RAM for R. In the process it tells you how much memory is currently being used by R.

There will likely be a tradeoff that you have to make in between size and runtime. For example, an algorithm that requires a lot of memory, may run faster, while a slower algorithm may require less memory. You will have to find the right balance between the two in order to provide a good experience to the user.
