---
title: "Data Science Capstone-Modeling"
output: html_document
author: M.Welt
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
library(tidyr)
```
```{r Cache_functions, echo=FALSE}
# Function to load saved R objects from a .RData file if the file exists
loadObjects <- function(fileName) {
  # Append ".RData" extension to the filename
  filePath <- file.path("cache", paste0(fileName, ".RData"))
  
  # Check if the file exists before attempting to load
  if (file.exists(filePath)) {
    # Load objects into the global environment
    load(filePath, envir = .GlobalEnv)
  }
}
save_objects <- function(objects, file_name) {
  if (!dir.exists("cache")) {
    dir.create("cache", recursive = TRUE)
  } 
  filePath <- file.path("cache", paste0(file_name, ".RData"))
  save(list = objects, file = filePath)
}
# Function to format numbers with thousand separators
format_number <- function(x) {
  if (is.numeric(x)) {
    format(x, big.mark = ",", scientific = FALSE)
  } else {
    x
  }
}

show_table <- function(report, rowNames = TRUE) {
  kable(report, align = rep("r", ncol(report)), row.names = rowNames) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  column_spec(1:ncol(report), border_right = "1px solid lightgray")
}
```

### Abstract
<div style="font-size: 12px;">The goal is to build first simple model for the relationship between words. 
This is the first step in building a predictive text mining application. 

**Tasks to accomplish**

 1. Build basic n-gram model - using the Exploratory Analysis(Previous exercise), build a basic 
n-gram model for predicting the next word based on the previous 1, 2, or 3 words.
 2. Build a model to handle unseen n-grams - in some cases people will want to type a combination of words that does not appear in the corpora.
 
**Questions to consider**

 1. How can you efficiently store an n-gram model (think Markov Chains)?
 2. How can you use the knowledge about word frequencies to make your model smaller and more efficient?
 3. How many parameters do you need (i.e. how big is n in your n-gram model)?
 4. Can you think of simple ways to "smooth" the probabilities (think about giving all n-grams a non-zero probability even if they aren't observed in the data) ?
 5. How do you evaluate whether your model is any good?
 6. How can you use [backoff models](https://en.wikipedia.org/wiki/Katz%27s_back-off_model) to estimate the probability of unobserved n-grams?
  7. Minimize size and runtime of the model in order to provide a reasonable experience to the user.

</div>

```{r, echo=FALSE}
# Tips, tricks, and hints
# 
# Here are a few tools that may be of use to you as you work on their algorithm:
# 
# object.size(): this function reports the number of bytes that an R object occupies in memory
# 
# Rprof(): this function runs the profiler in R that can be used to determine where bottlenecks in your function may exist. The profr package (available on CRAN) provides some additional tools for visualizing and summarizing profiling data.
# 
# gc(): this function runs the garbage collector to retrieve unused RAM for R. In the process it tells you how much memory is currently being used by R.
# 
# There will likely be a tradeoff that you have to make in between size and runtime. For example, an algorithm that requires a lot of memory, may run faster, while a slower algorithm may require less memory. You will have to find the right balance between the two in order to provide a good experience to the user.

# **Hints, tips, and tricks**
# 
# Two key aspects that we will have to keep in mind are the size and runtime of the algorithm. 
# 
#  - **Size**: the amount of memory (physical RAM) required to run the model in R
#  - **Runtime**: The amount of time the algorithm takes to make a prediction given the acceptable input
# 
# The goal for this prediction model is to minimize both the size and runtime of the model in order to provide a reasonable experience to the user.
```


### Task 1. Building a Basic N-Gram Model for Predictive Text
The goal of this task is to create an n-gram model that predicts the next word based on the previous 1, 2, or 3 words. 

#### Step 1: Tokenize text while preserving the sentence boundaries.

```{r, message=FALSE, warning=FALSE}
library(quanteda)

# Load text data
loadObjects("text_data")
# text_data <- head(text_data)  # Take first few lines
```
<div style="font-size: 10px;">
`r paste(head(text_data), collapse = "\n\n")``
</div>

```{r, message=FALSE, warning=FALSE}
text_data <- tolower(text_data)
text_data <- gsub("[.!?]", "✔", text_data)
text_data <- paste0(text_data, "✔")
# Join all text into a single string
text_data <- paste(text_data, collapse = " ")

# Remove punctuation during word tokenization
tokens_words <- tokens(text_data, what = "word", remove_punct = TRUE)  
```




#### Step 2: Generate bigrams- trigrams- and 4-gram
```{r Prepare parrallel process, message=FALSE}
library(dplyr)  
library(doParallel)
# Initialize parallel cluster
parallelCluster <- NULL

# Function to stop the parallel cluster
stopParallelCluster <- function() {
  if (!is.null(parallelCluster)) {
    stopCluster(parallelCluster)  # Stop the parallel cluster
    registerDoSEQ()  # Reset to sequential execution
    parallelCluster <<- NULL  # Clear the global cluster variable
  }
}

# Function to prepare the parallel cluster
prepareParallelCluster <- function() {
  num_cores <- max(detectCores() - 1, 1)  # Use all cores except one
  parallelCluster <<- makeCluster(num_cores)
  registerDoParallel(parallelCluster)
}

# Measure execution time


```

Parallel execution using foreach
```{r Generate bigrams- trigrams- and 4-gram, message=FALSE}
prepareParallelCluster()
start_time <- proc.time()
n_values <- c(2, 3, 4)  # Bigram, Trigram, Fourgram
ngrams_list <- vector("list", length(n_values))
ngrams_list <- foreach(i = seq_along(n_values), 
                        .combine = "c", 
                        .packages = "quanteda") %dopar% {
  result <- tokens_ngrams(tokens_words, n = n_values[i], concatenator = " ")
  if (length(result) == 0) list(NA) else as.list(result)
}
execution_time <- proc.time() - start_time
stopParallelCluster()
execution_time
```


```{r Retrieve n-grams, message=FALSE}

# Extract results
bigrams   <- unlist(ngrams_list[[1]], recursive = TRUE, use.names = FALSE)
trigrams  <- unlist(ngrams_list[[2]], recursive = TRUE, use.names = FALSE)
fourgrams <- unlist(ngrams_list[[3]], recursive = TRUE, use.names = FALSE)
```




#### Step 3: Build a Frequency Table
 - Count the occurrences of each n-gram
 - Store them in a sorted order (most frequent first)

 Convert to frequency table 
```{r  Convert to frequency table}
bigram_freq <- table(unlist(bigrams))
trigram_freq <- table(unlist(trigrams))
fourgram_freq <- table(unlist(fourgrams))
```

Convert to data frames and sort by frequency
```{r Convert to data frames and sort by frequency}
bigram_df <- as.data.frame(bigram_freq, stringsAsFactors = FALSE) %>%
  rename(ngram = Var1, frequency = Freq) %>% 
  filter(!grepl("✔", ngram, fixed = TRUE)) %>%
  separate(ngram, 
           into = c("word1", "word2"), sep = " ", remove = FALSE, extra = "merge")  %>%
  arrange(desc(frequency))


trigram_df <- as.data.frame(trigram_freq, stringsAsFactors = FALSE) %>%
  rename(ngram = Var1, frequency = Freq)  %>% 
  filter(!grepl("✔", ngram, fixed = TRUE))  %>%
  separate(ngram, 
           into = c("word1", "word2", "word3"), sep = " ", remove = FALSE, extra = "merge")  %>%
  select(-ngram) %>%
  arrange(desc(frequency))

fourgram_df <- as.data.frame(fourgram_freq, stringsAsFactors = FALSE) %>%
  rename(ngram = Var1, frequency = Freq)%>% 
  filter(!grepl("✔", ngram, fixed = TRUE)) %>%
  separate(ngram, 
           into = c("word1", "word2", "word3", "word4"), sep = " ", remove = FALSE, extra = "merge")  %>%
  select(-ngram) %>%
  arrange(desc(frequency))
```


```{r, echo=FALSE}
# Print the top n-grams
cat("Top Bigrams:")
print(head(bigram_df))

cat("Top Trigrams:")
print(head(trigram_df))

cat("Top 4-grams:")
print(head(fourgram_df))
```

We can observe that at text_data[6] begins with **"Parents"** and text_data[5] ends with **"you"**. 
Lets's check that no bigram is there like "you"-"parents". Thus the predictions will not cross the sentence boundaries.

```{r}
filtered_df <- bigram_df %>% 
  filter(grepl("parents", ngram, ignore.case = TRUE))

print(filtered_df)
```

#### Step 4: Prepare Data for Prediction

Convert the n-grams into a lookup table
Allow searching for most likely next words based on previous words


```{r}
bigram_predict_next_word <- function(word1, top_n = 3) {
  # Ensure input words are lowercase for consistency
  word1 <- tolower(word1)

  # Filter the 2gram table for matches
  suggestions <- bigram_df %>%
    filter(word1 == !!word1)
  
  # Check if any results are found before proceeding
  if (nrow(suggestions) == 0) {
    return("No suggestions found.")
  }
  
  # Select top predictions and extract the next word (word3)
  suggestions <- suggestions %>%
    arrange(desc(frequency)) %>%
    slice_head(n = top_n) %>%
    select(word2) %>%
    pull()

  return(suggestions)
}
trigram_predict_next_word <- function(word1, word2, top_n = 3) {
  # Ensure input words are lowercase for consistency
  word1 <- tolower(word1)
  word2 <- tolower(word2)
  
  # Filter the trigram table for matches
  suggestions <- trigram_df %>%
    filter(word1 == !!word1 & word2 == !!word2)
  
  # Check if any results are found before proceeding
  if (nrow(suggestions) == 0) {
    return("No suggestions found.")
  }
  
  # Select top predictions and extract the next word (word3)
  suggestions <- suggestions %>%
    arrange(desc(frequency)) %>%
    slice_head(n = top_n) %>%
    select(word3) %>%
    pull()

  return(suggestions)
}
fourgram_predict_next_word <- function(word1, word2, word3, top_n = 3) {
  # Ensure input words are lowercase for consistency
  word1 <- tolower(word1)
  word2 <- tolower(word2)
  word3 <- tolower(word3)
  
  # Filter the 4gram table for matches
  suggestions <- fourgram_df %>%
    filter(word1 == !!word1 & word2 == !!word2 & word3 == !!word3)
  
  # Check if any results are found before proceeding
  if (nrow(suggestions) == 0) {
    return("No suggestions found.")
  }
  
  # Select top predictions and extract the next word (word3)
  suggestions <- suggestions %>%
    arrange(desc(frequency)) %>%
    slice_head(n = top_n) %>%
    select(word4) %>%
    pull()

  return(suggestions)
}
```
Now let's suggest a predictions for trigram "a foreign":

```{r}
predicted_words <- trigram_predict_next_word("a", "foreign")
print(predicted_words)
```

### Task 2. Handle unseen n-gram
Build a model to handle unseen n-grams - in some cases people will want to type a combination of words that does not appear in the corpora.
To handle unseen n-grams, we can build a backoff model that:

 1. Uses trigrams if available
 2. Falls back to bigrams if no trigram is found
 3. Falls back to unigrams if no bigram is found
 4. Provides a default suggestion if nothing is found
 
 
```{r}
ngram_predict_next_word <- function(text, top_n = 3) {
  words <- unlist(strsplit(tolower(text), " "))
  word1 <- ifelse(length(words) >= 1, words[1], "")
  word2 <- ifelse(length(words) >= 2, words[2], "")
  word3 <- ifelse(length(words) >= 3, words[3], "")
  
  
  # Filter the 4gram table for matches
  suggestions <- fourgram_df %>%
    filter(word1 == !!word1 & word2 == !!word2 & word3 == !!word3)
  
  # Check if any results are found before proceeding
  if (nrow(suggestions) > 0) {
     suggestions <- suggestions %>%
      arrange(desc(frequency)) %>%
      slice_head(n = top_n) 
  
    return(suggestions)
  }
  
  word1 <- word2
  word2 <- word3
  # Filter the 3gram table for matches
  suggestions <- trigram_df %>%
    filter(word1 == !!word1 & word2 == !!word2)
  
  # Check if any results are found before proceeding
  if (nrow(suggestions) > 0) {
    # Select top predictions and extract the next word (word3)
    suggestions <- suggestions %>%
      arrange(desc(frequency)) %>%
      slice_head(n = top_n)
  
    return(suggestions)
    
  }
  
  word1 <- word2
  # Filter the 2gram table for matches
  suggestions <- bigram_df %>%
    filter(word1 == !!word1)
  
  # Check if any results are found before proceeding
  if (nrow(suggestions) > 0) {
    # Select top predictions and extract the next word (word3)
    suggestions <- suggestions %>%
      arrange(desc(frequency)) %>%
      slice_head(n = top_n)
  
    return(suggestions)
  }
  return("No suggestions found.")
}

ngram_predict_next_word("a case of")
```
 


```{r}

ngram_predict_next_word("a case of")
ngram_predict_next_word("would mean the")
ngram_predict_next_word("make me the")
ngram_predict_next_word("struggling but the")
ngram_predict_next_word("date at the")
ngram_predict_next_word("be on my")
ngram_predict_next_word("in quite some")
ngram_predict_next_word("with his little")
ngram_predict_next_word("faith during the")
ngram_predict_next_word("you must be")

```

