---
title: "Data Science Capstone-Modeling"
output: html_document
author: M.Welt
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(kableExtra)
```
```{r Cache_functions, echo=FALSE}
# Function to load saved R objects from a .RData file if the file exists
loadObjects <- function(fileName) {
  # Append ".RData" extension to the filename
  filePath <- file.path("cache", paste0(fileName, ".RData"))
  
  # Check if the file exists before attempting to load
  if (file.exists(filePath)) {
    # Load objects into the global environment
    load(filePath, envir = .GlobalEnv)
  }
}
save_objects <- function(objects, file_name) {
  if (!dir.exists("cache")) {
    dir.create("cache", recursive = TRUE)
  } 
  filePath <- file.path("cache", paste0(file_name, ".RData"))
  save(list = objects, file = filePath)
}
# Function to format numbers with thousand separators
format_number <- function(x) {
  if (is.numeric(x)) {
    format(x, big.mark = ",", scientific = FALSE)
  } else {
    x
  }
}

show_table <- function(report, rowNames = TRUE) {
  kable(report, align = rep("r", ncol(report)), row.names = rowNames) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) %>%
  column_spec(1:ncol(report), border_right = "1px solid lightgray")
}
```

### Abstract
<div style="font-size: 12px;">The goal is to build first simple model for the relationship between words. 
This is the first step in building a predictive text mining application. 

**Tasks to accomplish**

 1. Build basic n-gram model - using the exploratory analysis In Task2, build a basic 
n-gram model for predicting the next word based on the previous 1, 2, or 3 words.
 2. Build a model to handle unseen n-grams - in some cases people will want to type a combination of words that does not appear in the corpora.
 3.  Build a model to handle cases where a particular n-gram isn't observed.

**Questions to consider**

 1. How can you efficiently store an n-gram model (think Markov Chains)?
 2. How can you use the knowledge about word frequencies to make your model smaller and more efficient?
 3. How many parameters do you need (i.e. how big is n in your n-gram model)?
 4. Can you think of simple ways to "smooth" the probabilities (think about giving all n-grams a non-zero probability even if they aren't observed in the data) ?
 5. How do you evaluate whether your model is any good?
 6. How can you use [backoff models](https://en.wikipedia.org/wiki/Katz%27s_back-off_model) to estimate the probability of unobserved n-grams?
  7. Minimize size and runtime of the model in order to provide a reasonable experience to the user.

</div>

```{r, echo=FALSE}
# Tips, tricks, and hints
# 
# Here are a few tools that may be of use to you as you work on their algorithm:
# 
# object.size(): this function reports the number of bytes that an R object occupies in memory
# 
# Rprof(): this function runs the profiler in R that can be used to determine where bottlenecks in your function may exist. The profr package (available on CRAN) provides some additional tools for visualizing and summarizing profiling data.
# 
# gc(): this function runs the garbage collector to retrieve unused RAM for R. In the process it tells you how much memory is currently being used by R.
# 
# There will likely be a tradeoff that you have to make in between size and runtime. For example, an algorithm that requires a lot of memory, may run faster, while a slower algorithm may require less memory. You will have to find the right balance between the two in order to provide a good experience to the user.

# **Hints, tips, and tricks**
# 
# Two key aspects that we will have to keep in mind are the size and runtime of the algorithm. 
# 
#  - **Size**: the amount of memory (physical RAM) required to run the model in R
#  - **Runtime**: The amount of time the algorithm takes to make a prediction given the acceptable input
# 
# The goal for this prediction model is to minimize both the size and runtime of the model in order to provide a reasonable experience to the user.
```


### Task 1. Building a Basic N-Gram Model for Predictive Text
The goal of this task is to create an n-gram model that predicts the next word based on the previous 1, 2, or 3 words. 

#### Step 1: Tokenize text while preserving the sentence boundaries.

```{r, message=FALSE, warning=FALSE}
library(quanteda)

# Load text data
loadObjects("text_data")
text_data <- head(text_data)  # Take first few lines
# Fix missing sentence endings
text_data <- ifelse(grepl("[.!?]$", text_data), text_data, paste0(text_data, "."))
# Ensure each sentence has boundaries markers**
text_data <- paste("<s>", text_data, "</s>")  # Add markers
# Join all text into a single string (to preserve sentence structure)
text_data <- paste(text_data, collapse = " ")
# Tokenize sentences first (preserve sentence boundaries)
tokens_sentences <- tokens(text_data, what = "sentence")

# Convert sentence tokens into a character vector
sentences_text <- as.character(tokens_sentences)
# Tokenize words within each sentence
tokens_words <- tokens(sentences_text, what = "word", remove_punct = FALSE)

# Remove punctuation after word tokenization
tokens_clean <- tokens_remove(tokens_words, pattern = "[[:punct:]]", padding = FALSE)
# Convert all tokens to lowercase
tokens_clean <- tokens_tolower(tokens_clean)

# Convert the final cleaned tokens into a list for inspection
tokens_list <- as.list(tokens_clean)
# We keep punctuation so the model can predict sentence endings

tokens_df <- data.frame(tokens = unlist(tokens_clean))
head(tokens_df)
```


#### Step 2: Generate bigrams- trigrams- and 4-gram
```{r Generate bigrams- trigrams- and 4-gram, message=FALSE}
library(dplyr)  # Load the package

# 1️⃣ **Fix missing sentence endings**
text_data <- ifelse(grepl("[.!?]$", text_data), text_data, paste0(text_data, "."))

# Join all text into a single string
text_data <- paste(text_data, collapse = " ")

# 2️⃣ **Tokenize sentences first (preserve sentence boundaries)**
tokens_sentences <- tokens(text_data, what = "sentence")

# 3️⃣ **Remove punctuation from sentences**
tokens_sentences <- tokens_remove(tokens_sentences, pattern = "[[:punct:]]", padding = FALSE)

# 4️⃣ **Tokenize words within each sentence**
tokens_words <- tokens(text_data, what = "word", remove_punct = TRUE)  # Remove punctuation during word tokenization

# 5️⃣ **Generate n-grams**
bigrams <- tokens_ngrams(tokens_words, n = 2, concatenator = " ")
trigrams <- tokens_ngrams(tokens_words, n = 3, concatenator = " ")
fourgrams <- tokens_ngrams(tokens_words, n = 4, concatenator = " ")

# 6️⃣ **Convert to frequency tables**
bigram_freq <- table(unlist(bigrams))
trigram_freq <- table(unlist(trigrams))
fourgram_freq <- table(unlist(fourgrams))

# 7️⃣ **Convert to data frames and sort by frequency**
bigram_df <- as.data.frame(bigram_freq, stringsAsFactors = FALSE) %>%
  rename(ngram = Var1, frequency = Freq) 

trigram_df <- as.data.frame(trigram_freq, stringsAsFactors = FALSE) %>%
  rename(ngram = Var1, frequency = Freq)  

fourgram_df <- as.data.frame(fourgram_freq, stringsAsFactors = FALSE) %>%
  rename(ngram = Var1, frequency = Freq) 

# Print the top n-grams
cat("Top Bigrams:\n")
print(head(bigram_df, 10))

cat("\nTop Trigrams:\n")
print(head(trigram_df, 10))

cat("\nTop 4-grams:\n")
print(head(fourgram_df, 10))
```



Generate n-grams

Create bigrams, trigrams, and 4-grams
Preserve sentence boundaries to prevent incorrect word associations
3 Build a Frequency Table

Count the occurrences of each n-gram
Store them in a sorted order (most frequent first)
4 Prepare Data for Prediction

Convert the n-grams into a lookup table
Allow searching for most likely next words based on previous words



